---
title: "NFL Wide Receiver Combine Analysis"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
date: "2026-09-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,       # show code
  comment = "",       # drop the "##" prefix on printed output
  message = FALSE,    # hide package/import chatter (e.g. readr's column-spec messages)
  warning = FALSE      # hide non-fatal warnings from cluttering the report
)
```

## Overview

**Question:** Do NFL Scouting Combine metrics predict a wide receiver's career
success, measured as Approximate Value (AV) per season?

**Data sources:**

- `data/sportsref_download.xlsx` — career stats for wide receivers drafted
  2000 or later (Pro Football Reference / Stathead)
- `data/NFL_Combine_Since_2000.csv` — combine testing results, 2000–2025
  (Kaggle: `isaaksipple/nfl-combine-2000-2025`)

The analysis below loads and merges these two sources, visualizes each
combine metric against career AV per season, and fits a few linear
regression models to see which metrics (if any) are meaningful predictors.

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(readxl)
library(stringr)
```

## 1. Load data

Read in the raw career-stats and combine-testing files.

```{r load-data}
career_data <- read_excel("data/sportsref_download.xlsx")
combine_data <- read_csv("data/NFL_Combine_Since_2000.csv")
```

## 2. Filter to wide receivers

Both datasets cover every position, so we narrow each one down to wide
receivers. On the career side we also drop anyone drafted before 2000,
since that's the range the combine data covers.

```{r filter-wr}
combine_wr <- combine_data %>%
  filter(Position == "WR")

career_wr <- career_data %>%
  filter(Pos == "WR", From >= 2000)
```

## 3. Standardize player names for joining

Player names differ slightly between sources (periods, suffixes like
Jr./II/III), so we normalize both sides before joining on name + draft
year.

```{r clean-names}
clean_names <- function(x) {
  x %>%
    str_to_lower() %>%
    str_replace_all("\\.", "") %>%
    str_replace_all(" jr", "") %>%
    str_replace_all(" iii", "") %>%
    str_replace_all(" ii", "") %>%
    str_squish()
}

career_wr$Player <- clean_names(career_wr$Player)
combine_wr$Player <- clean_names(combine_wr$Player)
```

## 4. Merge career outcomes with combine testing data

Join on player name and draft year (`From` on the career side matches
`Year` on the combine side), then clean up the resulting column names and
derive `AV_per_season` — the outcome variable used throughout the rest of
the analysis — as career AV divided by number of seasons played.

```{r merge-data}
merged_data <- inner_join(
  career_wr, combine_wr,
  by = c("Player", "From" = "Year")
) %>%
  distinct() %>%
  rename(
    Career_AV      = `AV...3`,
    Forty          = `40-yd Dash`,
    Vertical_Jump  = `Vertical Jump`,
    Bench_Press    = `Bench Press`,
    Broad_Jump     = `Broad Jump`,
    Cone_Drill     = `3-Cone Drill`,
    Shuttle        = `20-yd Shuttle`,
    Draft_Team     = Team.y
  ) %>%
  select(-Team.x, -Rk, -Position, -AV...9) %>%
  mutate(
    Seasons       = To - From + 1,
    AV_per_season = Career_AV / Seasons
  )

combine_metrics <- c("Forty", "Vertical_Jump", "Bench_Press",
                      "Broad_Jump", "Cone_Drill", "Shuttle")

merged_data[combine_metrics] <- lapply(merged_data[combine_metrics], as.numeric)

cat("Merged dataset:", nrow(merged_data), "wide receivers\n")
```

## 5. Visualize each combine metric vs. career performance

For each combine metric we plot AV per season against that metric with a
linear trend line. Timed drills (40-yard dash, 3-cone, shuttle) have their
x-axis reversed, since a *lower* time is the better athletic outcome —
this keeps "better performance" pointing the same direction (right) across
every plot.

```{r save-plots}
combine_untimed_metrics <- c("Vertical_Jump", "Bench_Press", "Broad_Jump")
combine_timed_metrics   <- c("Forty", "Cone_Drill", "Shuttle")

dir.create("plots", showWarnings = FALSE)

for (metric in combine_untimed_metrics) {
  p <- ggplot(merged_data, aes(x = .data[[metric]], y = AV_per_season)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = paste(metric, "vs AV Per Season"), x = metric, y = "AV Per Season")
  ggsave(filename = paste0("plots/", metric, "_vs_AV.png"), plot = p, width = 8, height = 6)
}

for (metric in combine_timed_metrics) {
  # Lower (faster) times are better, so the x-axis is reversed for readability
  p <- ggplot(merged_data, aes(x = .data[[metric]], y = AV_per_season)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    scale_x_reverse() +
    labs(title = paste(metric, "vs AV Per Season"), x = metric, y = "AV Per Season")
  ggsave(filename = paste0("plots/", metric, "_vs_AV.png"), plot = p, width = 8, height = 6)
}
```

The loop above saves each plot as a PNG to `plots/` (useful for reuse
elsewhere), but doesn't display them inline in this report. The chunk below
renders each one directly in the knitted document so the plots and their
regression context sit together.

```{r show-plots, message=FALSE, warning=FALSE}
for (metric in c(combine_untimed_metrics, combine_timed_metrics)) {
  reverse_axis <- metric %in% combine_timed_metrics
  p <- ggplot(merged_data, aes(x = .data[[metric]], y = AV_per_season)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = paste(metric, "vs AV Per Season"), x = metric, y = "AV Per Season")
  if (reverse_axis) p <- p + scale_x_reverse()
  print(p)
}
```

## 6. Regression models

Three linear models test different subsets of combine metrics as
predictors of `AV_per_season`:

- **Full model** — all six combine metrics together
- **Speed model** — just the three timed drills (40-yard dash, shuttle,
  3-cone)
- **Explosion model** — just the two jump metrics (vertical, broad)

For each, `summary()` reports the coefficient estimates, their standard
errors, t-values, and p-values (is each metric individually a significant
predictor once the others are accounted for), along with the model's
overall R-squared (how much of the variance in AV per season the model
explains) and F-statistic (whether the model as a whole is a significant
predictor).

```{r regression-full}
full_model <- lm(
  AV_per_season ~ Forty + Vertical_Jump + Bench_Press + Broad_Jump + Cone_Drill + Shuttle,
  data = merged_data
)
summary(full_model)
```

```{r regression-speed}
speed_model <- lm(
  AV_per_season ~ Forty + Shuttle + Cone_Drill,
  data = merged_data
)
summary(speed_model)
```

```{r regression-explosion}
explosion_model <- lm(
  AV_per_season ~ Vertical_Jump + Broad_Jump,
  data = merged_data
)
summary(explosion_model)
```

**Reading these results:** look at each model's R-squared to see how much
of the spread in career AV per season it accounts for, and check which
individual predictors have p-values below 0.05 (conventionally
"significant") to see whether any single combine metric stands out. Given
these are raw combine numbers predicting a noisy, multi-year career outcome,
don't expect a very high R-squared — the interesting part is usually the
*relative* pattern across the three models (e.g., does the speed model or
the explosion model explain more, and does any individual metric hold up
once the others are controlled for).

## 7. Combined explosion score

As a follow-up, we combine vertical jump and broad jump into a single
"Explosion Score" and plot it against career AV per season.

```{r explosion-score}
merged_data <- merged_data %>%
  mutate(Explosion_Score = Vertical_Jump + Broad_Jump)

explosion_plot <- ggplot(merged_data, aes(x = Explosion_Score, y = AV_per_season)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(
    title = "Explosion Score (Vertical + Broad Jump) vs AV Per Season",
    x = "Explosion Score",
    y = "AV Per Season"
  )
ggsave("plots/Explosion_Score_vs_AV.png", plot = explosion_plot, width = 8, height = 6)

explosion_plot
```

## Summary

This report walked through: merging combine testing data with career
outcomes for WRs drafted since 2000, visualizing each combine metric
against career AV per season, and fitting three regression models (full,
speed-only, explosion-only) to test which combine metrics best predict
career success. The regression tables above are the core results — refer
to each model's R-squared and per-metric p-values to judge which combine
metrics, if any, carry predictive signal for WR career performance.
